SmartCheck

### 2012-11-12 paper/paper.tex
paper updates.

The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Nowadays, QuickCheck is a standard
stage of a program development cycle in Haskell.

Because QuickCheck generates random values for testing, counterexamples it finds
may not be minimal.  In their original QuickCheck paper~\cite{qc}, the authors
report the following user experience by Andy Gill:
%
\begin{quote}
Sometimes the counter examples found are very large and it is difficult to go
back to the property and understand why it is a counter example.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that defines a function
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a] 
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

One of the dirty little secrets of QuickCheck is that even for simple data types
and simple properties, counterexamples can be much larger than the minimal
counterexamples.  Intuitively, there are two reasons: (1) because values are
generated randomly, a smaller value that produces a counterexample is skipped;
and (2) for a composite type, larger values are generated in tandem.  For
example, consider the following data type and property:
%
\begin{code}
data A = A Int deriving Show

@@

prop :: [A] -> Property
prop x = property $ length x < 5
\end{code}
%
\noindent
where \ttp{property} is a QuickCheck library function that lifts a \ttp{Bool} into
the \ttp{Property} type.  (We wrap \ttp{Int} in a constructor to prevent QuickCheck
from using a default implementation of \ttp{shrink} for \ttp{Int}.)  A typical
counterexample returned by QuickCheck is
%
\begin{code}
> [A (-2),A (-9),A (-26),A (-28),A (-22)]  
\end{code}
%
\noindent
Whereas the smallest counterexample is
%
\begin{code}
> [A 0,A 0,A 0,A 0,A 0]
\end{code}
%
\noindent
The problem is that while QuickCheck is generating larger lists, it is also
generating larger integers.

SmallCheck is another testing framework that is guaranteed to return the a
smallest counterexample, if one exists~\cite{sc}.  SmallCheck does this by
enumerating all possible inputs, ordered from smallest to largest, up to some
user-defined bound.  While SmallCheck is effective for testing many programs and
properties (in accordance with the small model principle~\cite{jackson}),
counterexamples deep in the state space may be infeasible to uncover.  For
example, consider again our data type \ttp{A} defined above, used in the
following property:
%
\begin{code}
data B = B A A A A deriving Show

prop :: B -> Bool
prop (B a b c d) = and $ map f [a, b, c, d]
  where f (A x) = x < 100
\end{code}
%
\noindent
With SmallCheck, a property is checked up to a depth set by the user, measuring
how large values are.  For example, for type \ttp{Int}, the depth \ttp{d} is
defined as the integers in the list \ttp{(-d)..d]}.  For a product type,
  SmallCheck generates the cross product of its fields.  In the example above,
  SmallCheck (or Lazy SmallCheck) will find a counterexample at depth 101,
  requiring approximately $100^4$ test cases, which is infeasible to compute.\footnote{This
  is an upper-bound, since Lazy SmallCheck eliminates some redundant test-cases,
  but computing a counterexample is still infeasible.}

>>>>

The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Nowadays, QuickCheck is a standard
stage of a program development cycle in Haskell.  Because QuickCheck generates
random values for testing, counterexamples it finds may not be substantially
larger than a minimal counterexample.  In their original QuickCheck
paper~\cite{qc}, the authors report the following user experience by Andy Gill:
%
\begin{quote}
Sometimes the counter examples found are very large and it is difficult to go
back to the property and understand why it is a counter example.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that defines a function
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a] 
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

This trouble with randomly generating large counterexamples is compounded in
composite data types, for which increasingly larger sub-values are generated in
tandem.  For example, consider the following data type and property:
%
\begin{code}
data A = A Int deriving Show

@@

prop :: [A] -> Property
prop x = property (length x < 5)
\end{code}
%
\noindent
where \ttp{property} is a QuickCheck library function that lifts a \ttp{Bool} into
the \ttp{Property} type.  (We wrap \ttp{Int} in a constructor to prevent QuickCheck
from using a default implementation of \ttp{shrink} for \ttp{Int}.)  A typical
counterexample returned by QuickCheck is
%
\begin{code}
> [A (-2),A (-9),A (-26),A (-28),A (-22)]  
\end{code}
%
\noindent
Whereas the smallest counterexample is
%
\begin{code}
> [A 0,A 0,A 0,A 0,A 0]
\end{code}
%
\noindent
The problem is that while QuickCheck is generating larger lists, it is also
generating larger integers.

SmallCheck is another testing framework that is guaranteed to return the a
smallest counterexample, if one exists~\cite{sc}.  SmallCheck does this by
enumerating all possible inputs, ordered from smallest to largest, up to some
user-defined bound.  While SmallCheck is effective for testing many programs and
properties (in accordance with the small model principle~\cite{jackson}),
counterexamples to even relatively simple properties may be infeasible to
uncover for product types.  For
example, consider again our data type \ttp{A} defined above, used in the
following property:
%
\begin{code}
data B = B A A A A deriving Show

prop :: B -> Bool
prop (B a b c d) = and $ map f [a, b, c, d]
  where f (A x) = x < 100
\end{code}
%
\noindent
With SmallCheck, a property is checked up to a depth set by the user, measuring
how large values are.  For example, for type \ttp{Int}, the depth \ttp{d} is
defined as the integers in the list \ttp{(-d)..d]}.  For a product type,
  SmallCheck generates the cross product of its fields.  In the example above,
  SmallCheck (or Lazy SmallCheck) will find a counterexample at depth 101,
  requiring on the order of $200^4$ test cases to be executed; executing $10e^6$
  tests requires just under four minutes on a Pentium i7 at
  2.7GHz.\footnote{This is an upper-bound, since Lazy SmallCheck eliminates some
    redundant test-cases.}

Furthermore, even if a minimal counterexample is found, regardless of the testing
framework used, it tells the user nothing about related counterexamples.  That
is, it would be nice for a counterexample to be generalized so that the user is
presented with a formula characterizing a set of counterexamples.

For
example,
